Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-11 manuscript/draft.Rmd
update model descriptions

Observer effects are inherent in large data sets taken by different observers,
and are known to occur in BBS [@sauer1994]. We estimated the observer effects,
and associated uncertainty about those effects, with a linear mixed model with
observer as a random effect. Because observer and site are strongly related (an
observer tends to repeatedly sample the same site), site was also included as a
random effect to ensure that inferred deviations were actually observer-related
(as opposed to being related to the sites that a given observer happened to
see). The resulting model partitions the variance in observed richness values
into site-level variance, observer-level variance, and residual variance
(e.g. variation within a site from year to year), and can therefore also be used
directly as the "average" baseline model (see below). The estimated observer
effects can be subtracted from the richness values for a particular observer to
provide an estimate of how many species would have been found by a "typical"
observer. To incorporate uncertainty in these "corrected" richness values into
the forecasting models we collected [[N]] Monte Carlo samples from the model's
posterior distribution, using Stan, and fit each of the downstream models with
each of the Monte Carlo samples.

@@

We fit three sets of models that were fit to each site independently. These
models were fit to each BBS route both on the raw richness values and on the
residuals from our observer model. When correcting for observer effects, we
averaged across [[N]] models that were fit separately to the [[N]] Monte Carlo
estimates of the observer effects, to account for our uncertainty in the true
values of those effects. All of these models use a Gaussian error distribution.

**Baseline models.** We used two simple models as baselines for comparison to
the fits of more complex models. These baselines treated site-level richness
observations either as uncorrelated noise around a site-level constant (the
"average" model) or as an autoregressive model with a single year of history
(the "naive" model) [@hyndman2014]. Predictions from the "average" model are
centered on the average richness and the confidence intervals are narrow and
constant-width. The "naive" model, in contrast, predicts that future
observations will be similar to the final observed value (e.g. the value
observed in `$r settings$last_train_year`), and the confidence intervals expand
rapidly as the predictions extend farther into the future.

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @hyndman2016] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components, and how many terms to include for each one,
using AICc [@hyndman2016]. Since there is no seasonable component to the BBS
time-series we did not include a season component in these models. Otherwise we
used the default settings for this function. [[Note: should probably record
somewhere what order model the auto.arima function tends to choose.  Eyeballing
the results, seems to be most similar to the naive model, but would be good to
check.]] We also fit an "Auto-ARIMA + environment" model, in which each
site-level time series could also be predicted by annual changes in our
environmental variables at that site.

@@

In contrast to the time-series models, most attempts to predict species richness
focusing on using correlative models to predict richness changes over time from
environmental data. We tested three common variants of this approach - direct
modeling of species richness, stacking individual species distribution models,
and joint species distribution models. Following the standard approach,
site-level random effects were not included in these models as predictors,
meaning that this approach implicitly assumes that two sites with identical
Bioclim, elevation, and NDVI values should have identical richness
distributions. As above, we included observer effects & the associated
uncertainty by running these models [[N]] times (once per MCMC sample).

**"Macroecological" model.** We used a generalized boosted regression tree model
using the `gbm` package [@ridgeway2017] to directly model species richness as a function of
environmental variables. Boosted regression trees are a form of tree based
modeling that work by fitting thousands of small tree-structured models
sequentially, with each tree optimized to reduce the error of its
predecessors. They are flexible models that are considered well suited for
prediction [@elith2008]. This model was optimized using a Gaussian
likelihood. [[other gbm settings go here.]]

**Species Distribution Models.** Species distribution models (SDMs) model
individual species' occurrence probabilities using environmental
variables. Species level models are used to predict richness by summing the
predicted probability of occupancy across all species at a site. We fit SDMs
using random forests, a common approach in the species distribution modeling
literature [[refs; would be good to provide a source that says these are a good
choice]]. Predicted probabilities of occurrence were summed directly to predict
the number of species. This avoids known problems with the use of thresholds for
determining whether or not a species will be present at a site [[refs]]. When
incorporating observer effects we avoided fitting [[N x S]] models ([[N]] MCMC
samples * [[S]] species) by fitting one regression tree using each of the Monte
Carlo estimates of our observer effects and averaging the predictions of the
resulting forest [[citation?]]. We used Gaussian approximations for uncertainty,
based on the means and variances of the sums of independent Bernoulli random
variables [[cite Calabrese et al.]].

**Joint Species Distribution Models.** Joint species distribution models (JSDMs)
are a new approach that makes predictions about the full composition of a
community instead of modeling each species independently. JSDMs remove the
assumed independence among species and explicitly account for the possibility
that a site will be much more (or less) suitable for birds in general, or
particular groups of birds, than one would expect based on the available
environmental measurements alone. As a result, JSDMs do a better job of
representing our uncertainty about richness, while stacked SDMs underestimate it
[@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

>>>>

Observer effects are inherent in large data sets taken by different observers,
and are known to occur in BBS [@sauer1994]. We estimated the observer effects
(and associated uncertainty about those effects) with a linear mixed model with
observer as a random effect built in the Stan probabilistic programming language
[@stan] and its R interface [@rstan]. Because observer and site are strongly
related (an observer tends to repeatedly sample the same site), site was also
included as a random effect to ensure that inferred deviations were actually
observer-related (as opposed to being related to the sites that a given observer
happened to see). The resulting model partitions the variance in observed
richness values into site-level variance, observer-level variance, and residual
variance (e.g. variation within a site from year to year), and can therefore
also be used directly as the "average" baseline model (see below). The estimated
observer effects can be subtracted from the richness values for a particular
observer to provide an estimate of how many species would have been found by a
"typical" observer. To incorporate uncertainty in these "corrected" richness
values into the forecasting models we collected 500 Monte Carlo samples from the
model's posterior distribution, and fit each of the downstream models with each
of the Monte Carlo samples.

@@

We fit three sets of models that were fit to each site independently. These
models were fit to each BBS route both on the raw richness values and on the
residuals from our observer model. When correcting for observer effects, we
averaged across 500 models that were fit separately to the 500 Monte Carlo
estimates of the observer effects, to account for our uncertainty in the true
values of those effects. All of these models use a Gaussian error distribution.

**Baseline models.** We used two simple models as baselines for comparison to
the fits of more complex models. These baselines treated site-level richness
observations either as uncorrelated noise around a site-level constant (the
"average" model) or as an autoregressive model with a single year of history
(the "naive" model) [@hyndman2014]. Predictions from the "average" model are
centered on the average richness observed during training, and the confidence
intervals are narrow and constant-width. The "naive" model, in contrast,
predicts that future observations will be similar to the final observed value
(e.g. the value observed in `$r settings$last_train_year`), and the confidence
intervals expand rapidly as the predictions extend farther into the future.

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @hyndman2016] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components, and how many terms to include for each one,
using AICc [@hyndman2016]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function. [[Note: should probably record
somewhere what order model the auto.arima function tends to choose.  Eyeballing
the results, seems to be most similar to the naive model, but would be good to
check.]]

@@

In contrast to the time-series models, most attempts to predict species richness
focusing on using correlative models to predict richness changes over time from
environmental data. We tested three common variants of this approach - direct
modeling of species richness, stacking individual species distribution models,
and joint species distribution models (JSDMs). Following the standard approach,
site-level random effects were not included in these models as predictors,
meaning that this approach implicitly assumes that two sites with identical
Bioclim, elevation, and NDVI values should have identical richness
distributions. As above, we included observer effects & the associated
uncertainty by running these models 500 times (once per MCMC sample).

**"Macroecological" model.** We used a boosted regression tree model using the
`gbm` package [@ridgeway2017] to directly model species richness as a function
of environmental variables. Boosted regression trees are a form of tree based
modeling that work by fitting thousands of small tree-structured models
sequentially, with each tree optimized to reduce the error of its predecessors.
They are flexible models that are considered well suited for prediction
[@elith2008]. This model was optimized using a Gaussian likelihood, with a 
maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees.
The number of trees was selected using the "out of bag" estimator, and 
[[summarize how many trees were typically chosen here, for both observer-based
and observer-free models]].

**Species Distribution Models.** Species distribution models (SDMs) predict
individual species' occurrence probabilities using environmental variables.
Species-level models are used to predict richness by summing the predicted
probability of occupancy across all species at a site; following [[Calabrese et
al.]], we calculated the uncertainty in our richness estimate by treating
richness as a sum over independent Bernoulli random variables (one per species).
This avoids known problems with the use of thresholds for determining whether or
not a species will be present at a site [[refs]]. We represented SDMs in this
analysis using random forests, a common approach in the species distribution
modeling literature based on fitting hundreds of independent regression trees to
randomly-perturbed versions of the data [[refs; would be good to provide a
source that says these are a good choice]]. Each of the 500 trees in our
species-level random forests used a different Monte Carlo estimate of the
observer effects as a predictor variable.

**Joint Species Distribution Models.** Joint species distribution models (JSDMs)
are a new approach that makes predictions about the full composition of a
community instead of modeling each species independently as above. JSDMs remove
the assumed independence among species and explicitly account for the
possibility that a site will be much more (or less) suitable for birds in
general, or particular groups of birds, than one would expect based on the
available environmental measurements alone. As a result, JSDMs do a better job
of representing our uncertainty about richness, while stacked SDMs underestimate
it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.
